---
title: "MSc Dissertation"
author: '220225605'
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

# MSc Dissertation  

## Does ethnicity and income moderate the relationship between the consumption of saturated fats and free sugars and BMI?  

### Loading the required libraries  

```{r setup, echo = TRUE, warning = FALSE}
library(here)
library(tidyverse)
library(naniar)
```

### Importing required datasets (Waves 1 to 4)

```{r import1, echo = TRUE, warning = FALSE}
# opening person level dietary records for waves 1 to 4
PLDR14 <- read_csv(here("data","PersonLevel1-4.csv"))

# retaining required columns from the data-set above 
PLDR14 <- PLDR14 %>% select(seriali, Ndays, Saturatedfattyacidsg, FreeSugarsg, Energykcal) 

# opening individual level data for waves 1 to 4 
Indiv14 <- read_csv(here("data", "Indiv1-4.csv"))

# retaining required columns from the data-set above 
Indiv14 <- Indiv14 %>% select(Sex, age, eqvinc, EthGrG, qual7, cigst2, NumCh118, seriali, bmival, CutDown, DMHSize)

# renaming a column so that the datasets (PLDR14 & Indiv14) can be merged 
Indiv14 <- Indiv14 %>% rename(EthGrpGB = EthGrG)

# merging the two datasets together
Data14 <- merge(x = Indiv14, y = PLDR14, by = "seriali", all.x = TRUE)
```

### Importing required datasets (Waves 5 to 6)

```{r import2, echo = TRUE, warning = FALSE}
# opening person level dietary records for waves 5 to 6
PLDR56 <- read_csv(here("data","PersonLevel5-6.csv"))

# retaining required columns from the data-set above 
PLDR56 <- PLDR56 %>% select(seriali, Ndays, Saturatedfattyacidsg, FreeSugarsg, Energykcal)

# opening individual level data for waves 5 to 6 
Indiv56 <- read_csv(here("data", "Indiv5-6.csv"))

# retaining required columns from the data-set above 
Indiv56 <- Indiv56 %>% select(Sex, age, eqvinc, EthGrpGB, qual7, cigst2, NumCh118, seriali, bmival, CutDown, DMHSize)

# merging the two datasets together
Data56 <- merge(x = Indiv56, y = PLDR56, by = "seriali", all.x = TRUE)
```

### Merging data-sets from 2008/2009 - 2013/2014

```{r merge, echo = TRUE, warning = FALSE}
FinalData <- rbind(Data14, Data56)
```

### Removing respondents below the age of 18 

```{r adults, echo = TRUE, warning = FALSE}
FinalData <- FinalData[FinalData$age > 17,]
```

### Dealing with missing values 

```{r missing, echo = TRUE, warning = FALSE}
# removing rows with the following values: -1 (not applicable), -4 (not applicable to survey year), -8 (Don't know/ Can't say), -9 (Refusal)
FinalData <- FinalData %>% replace_with_na(replace = list(bmival = c(-1, -4, -8, -9), Saturatedfattyacidsg = c(-1,-4, -8, -9), FreeSugarsg = c(-1, -4, -8, -9), eqvinc = c(-1,-4, -8, -9), cigst2 = c(-1,-4, -8, -9), NumCh118 = c(-1,-4, -8, -9), qual7 = c(-1,-4, -8, -9), Energykcal = c(-1,-4, -8, -9), DMHSize = c(-1,-4, -8, -9), CutDown = c(-1,-4, -8, -9), EthGrpGB = c(-1, -4, -8, -9)))
# drop rows with missing values 
FinalData <- FinalData %>% drop_na()
```

### Recoding the ethnicity column 

```{r ethnicity, echo = TRUE, warning = FALSE}
# white
FinalData$EthGrpGB[FinalData$EthGrpGB == 2] <- 1
FinalData$EthGrpGB[FinalData$EthGrpGB == 3] <- 1
FinalData$EthGrpGB[FinalData$EthGrpGB == 4] <- 1
# south asian 
FinalData$EthGrpGB[FinalData$EthGrpGB == 9] <- 2
FinalData$EthGrpGB[FinalData$EthGrpGB == 10] <- 2
FinalData$EthGrpGB[FinalData$EthGrpGB == 11] <- 2
# black 
FinalData$EthGrpGB[FinalData$EthGrpGB == 13] <- 3
FinalData$EthGrpGB[FinalData$EthGrpGB == 14] <- 3
FinalData$EthGrpGB[FinalData$EthGrpGB == 15] <- 3

# chinese 
FinalData$EthGrpGB[FinalData$EthGrpGB == 16] <- 4

# arab 
FinalData$EthGrpGB[FinalData$EthGrpGB == 17] <- 5

# dropping rows for mixed and 'any other' ethnicity groups 
FinalData <- FinalData %>% replace_with_na(replace = list(EthGrpGB = c(5, 6, 7, 8, 12, 18)))
FinalData <- FinalData %>% drop_na()
```

### Labelling factor variables 

```{r factor, echo = TRUE, warning = FALSE}
# converting numeric variables to factor variables 
# sex 
FinalData$Sex <-  cut(
  FinalData$Sex, 2, labels = c(
    'Male', 
    'Female'))
 
# number of children (converting a continuous variable to a binary, categorical variable)
FinalData$NumCh118[FinalData$NumCh118 > 0] <- "Have Children"
FinalData$NumCh118[FinalData$NumCh118 == 0] <- "Do not have Children"
FinalData$NumCh118 <- as.factor(FinalData$NumCh118)

# cigarette smoking status (converting a polytomous categorical variable to a binary, categorical variable)
FinalData$cigst2[FinalData$cigst2 == 5] <- "Non-smoker"
FinalData$cigst2[FinalData$cigst2 < 5] <- "Smoker"
FinalData$cigst2 <- as.factor(FinalData$cigst2)

# educational qualifications (converting a polytomous categorical variable to a binary, categorical variable)
FinalData$qual7[FinalData$qual7 <= 2] <- "Higher education and above"
FinalData$qual7[FinalData$qual7 == 3] <- "A level or equivalent"
FinalData$qual7[FinalData$qual7 == 4] <- "GCSE or equivalent"
FinalData$qual7[FinalData$qual7 == 5] <- "GCSE or equivalent"
FinalData$qual7[FinalData$qual7 == 6] <- "Foreign or other qualifications"
FinalData$qual7[FinalData$qual7 == 7] <- "No qualifications"
FinalData$qual7[FinalData$qual7 == 8] <- "Still in FT education"
FinalData$qual7 <- as.factor(FinalData$qual7)
```

### Categorise the outcome variable bmival for the logistic regression model 

```{r obese, echo = TRUE, warning = FALSE}
# using a single bmi cutoff point to categorise obesity
FinalData <- FinalData %>% mutate(bmi_uni = case_when(bmival >= 30 ~ "Obese", bmival < 30 ~ "Not Obese"))

# using multiple bmi cutoffs to categorise obesity on the basis of ethnicity 
FinalData <- FinalData %>% mutate(bmi_eth = case_when(EthGrpGB = 1 & bmival >=30 ~ 'Obese', EthGrpGB = 1 & bmival < 30 ~ "Not Obese", EthGrpGB = 2 & bmival >=24 ~ 'Obese', EthGrpGB = 2 & bmival < 24 ~ "Not Obese", EthGrpGB = 3 & bmival >=28 ~ 'Obese', EthGrpGB = 3 & bmival < 28 ~ "Not Obese", EthGrpGB = 4 & bmival >=27 ~ 'Obese', EthGrpGB = 4 & bmival < 27 ~ "Not Obese"))

# converting character/numeric variables into factor variables      
FinalData$bmi_uni <- as.factor(FinalData$bmi_uni)
FinalData$bmi_eth <- as.factor(FinalData$bmi_eth)
FinalData$EthGrpGB <- as.factor(FinalData$EthGrpGB)
```

### Saving the processed data 
```{r save, echo = TRUE, warning = FALSE}
# save file
save(FinalData, file = "processed data/ProcessedData.RData")
save(FinalData, file = "processed data/ProcessedData.csv")
```
 


