---
title: "Analysis"
author: '220225605'
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

# MSc Dissertation  

## Does ethnicity and income moderate the relationship between the consumption of saturated fats and free sugars and BMI?  

### Loading the required libraries  

```{r setup, echo = TRUE, warning = FALSE}
library(here)
library(tidyverse)
library(ggfortify)
```

### Importing cleaned and processed data 

```{r load, echo = TRUE, warning = FALSE}
load(here("processed data/ProcessedData.RData"))
```

### Visualising the outcome variable 

```{r viz_outcome, echo = TRUE, warning = FALSE}
# density plots for the outcome variable 
p <- ggplot(FinalData, aes(x = bmival))
p + geom_density(color = "darkblue", fill = "lightblue")

# checking for outliers
boxplot(FinalData$bmival)

# following the exclusion criteria (removal of participants with a BMI less than 18 and BMI above 45)
FinalDataTemp <- FinalData[FinalData$bmival > 18,]
FinalDataTemp <- FinalDataTemp[FinalDataTemp$bmival < 40,]

# checking for the range 
range(FinalDataTemp$bmival)

# density plot of the outcome variable after the removal of underweight and morbidly obese participants  
p1 <- ggplot(FinalDataTemp, aes(x = bmival))
p1 + geom_density(color = "darkblue", fill = "lightblue")
```

### Checking the relationship between predictors and the outcome variable 

```{r relationship, echo = TRUE, warning = FALSE}
# Predictor 1 - Saturated Fatty Acids ----------------------------------------
ggplot(FinalData, aes(x = Saturatedfattyacidsg)) + geom_density(color = "darkblue", fill = "lightblue")+ labs(x = "Saturated Fatty Acids",  y = "BMI")

# checking for outliers 
boxplot(FinalData$Saturatedfattyacidsg)

# removing outliers 
FinalDataTemp <- FinalDataTemp[FinalDataTemp$Saturatedfattyacidsg < 50,]
ggplot(FinalDataTemp, aes(x = Saturatedfattyacidsg)) + geom_density(color = "darkblue", fill = "lightblue")

# scatter plot with saturated fat intake on the x axis and BMI on the y axis 
p1 <- ggplot(FinalDataTemp, aes(x = Saturatedfattyacidsg, y = bmival))
p1 + geom_point(color = "grey0")+ labs(x = "Saturated Fatty Acids",  y = "BMI") + geom_smooth(method = "lm")

# Predictor 2 - Free Sugar Consumption-----------------------
ggplot(FinalData, aes(x = FreeSugarsg)) + geom_density(color = "darkblue", fill = "lightblue") + labs(x = "Free Sugars",  y = "BMI")

# checking for outliers 
boxplot(FinalData$FreeSugarsg)

# removal of outliers 
FinalDataTemp <- FinalDataTemp[FinalDataTemp$FreeSugarsg < 150,]
range(FinalDataTemp$FreeSugarsg)
ggplot(FinalDataTemp, aes(x = FreeSugarsg)) + geom_density(color = "darkblue", fill = "lightblue")

# scatter plot with free sugar intake on the x axis and BMI on the y axis
p2 <- ggplot(FinalDataTemp, aes(x = FreeSugarsg, y = bmival))
p2 + geom_point(color = "grey0")+ labs(x = "Free Sugars",  y = "BMI") + geom_smooth(method = "lm")
```

### Descriptive Statistics 

```{r desc, echo=TRUE, warning=FALSE}
# frequency table - weight statuses 
table(FinalDataTemp$bmi_eth)

# mean and standard deviation of age across weight statuses 
FinalDataTemp %>% group_by(bmi_eth) %>% summarise_at(vars(age, eqvinc), list(mean = mean, sd = sd))
sd(FinalDataTemp$age) # standard deviation of age 
# performing a t-test to compare the mean age across weight statuses 
boxplot(FinalDataTemp$age ~ FinalDataTemp$bmi_eth)
t.test (age ~ bmi_eth, var.equal=TRUE, data = FinalDataTemp)

# ethnicity across weight status 
table(FinalDataTemp$EthGrpGB, FinalDataTemp$bmi_eth)

# converting equivalised household income into three categories (tertiles)
FinalDataTemp <- FinalDataTemp %>%
  mutate(tertiles = ntile(eqvinc, 3)) %>%
  mutate(tertiles = if_else(tertiles == 1, 'Low', if_else(tertiles == 2, 'Medium', 'High')))
# equivalised income across weight status 
table(FinalDataTemp$tertiles, FinalDataTemp$bmi_eth)
# chi-square test (Categorical variables - income tertiles and weight status)
chisq.test(FinalDataTemp$tertiles, FinalDataTemp$bmi_eth, correct = FALSE)

# gender across weight statuses 
table(FinalDataTemp$Sex, FinalDataTemp$bmi_eth)
# chi-square test (Categorical variables - gender and weight status)
chisq.test(FinalDataTemp$Sex, FinalDataTemp$bmi_eth, correct = FALSE)

# qualifications across weight statuses 
table(FinalDataTemp$qual7, FinalDataTemp$bmi_eth)
# chi-square test (Categorical variables - educational qualifications and weight status)
chisq.test(FinalDataTemp$qual7, FinalDataTemp$bmi_eth, correct = FALSE)

# presence of children across weight statuses 
table(FinalDataTemp$NumCh118, FinalDataTemp$bmi_eth)
# chi-square test (Categorical variables - presence of children and weight status)
chisq.test(FinalDataTemp$NumCh118, FinalDataTemp$bmi_eth, correct = FALSE)

# current cigarette smoking status across weight statuses 
table(FinalDataTemp$cigst2, FinalDataTemp$bmi_eth)
# chi-square test (Categorical variables - cig. smoking status and weight status)
chisq.test(FinalDataTemp$cigst2, FinalDataTemp$bmi_eth, correct = FALSE)

# saturated fat intake status across weight status 
# calculating daily caloric intake from saturated fats alone 
FinalDataTemp <- FinalDataTemp %>% mutate(SF_calories = Saturatedfattyacidsg*9)
# creating a new column to determine if a calories from saturated fat exceed 10% of total caloric intake
FinalDataTemp <- FinalDataTemp %>% mutate(SF_Status = case_when(SF_calories >= 0.10*Energykcal ~ "Yes", SF_calories < 0.10*Energykcal ~ "No"))
# frequency table (Saturated Fat Intake and Weight Status)
table(FinalDataTemp$SF_Status, FinalDataTemp$bmi_eth)
# chi-square test (Categorical variables - saturated fat intake status and weight status)
chisq.test(FinalDataTemp$SF_Status, FinalDataTemp$bmi_eth, correct = FALSE)

# free sugar intake status across weight status 
# calculating daily caloric intake from free sugars alone 
FinalDataTemp <- FinalDataTemp %>% mutate(FS_calories = FreeSugarsg*4)
# creating a new column to determine if a calories from free sugars exceed 10% of total caloric intake
FinalDataTemp <- FinalDataTemp %>% mutate(FS_Status = case_when(FS_calories >= 0.10*Energykcal ~ "Yes", FS_calories < 0.10*Energykcal ~ "No"))
# frequency table (Free Sugar Intake and Weight Status)
table(FinalDataTemp$FS_Status, FinalDataTemp$bmi_eth)
# chi-square test (Categorical variables - free sugar intake status and weight status)
chisq.test(FinalDataTemp$FS_Status, FinalDataTemp$bmi_eth, correct = FALSE)

```


### Multiple Linear Regression 

#### Adding covariates to the model 

Including age, gender, educational attainments, presence of children and cigarette smoking status as predictors and BMI as the outcome variable 

```{r covariates, echo = TRUE, warning = FALSE}
FinalDataTemp$qual7 <- relevel(FinalDataTemp$qual7, ref = "No qualifications")
lm1 <- lm(bmival ~ age + Sex + qual7 + NumCh118 +cigst2, data = FinalDataTemp)
summary(lm1)
# confidence intervals 
confint(lm1,level = 0.95)
```

#### Adding the primary predictors to the model 

Including saturated fatty acids intake and free sugars intake to the model controlling for the covariates 

```{r lm2, echo = TRUE, warning = FALSE}
FinalDataTemp$qual7 <- relevel(FinalDataTemp$qual7, ref = "No qualifications")
lm2 <- lm(bmival ~ age + Sex + qual7 + NumCh118 +cigst2 +Saturatedfattyacidsg + FreeSugarsg, data = FinalDataTemp)
summary(lm2)
confint(lm2, level = 0.95)
```

#### Including ethnicity and income as possible predictors (main effects)

```{r lm3, echo = TRUE, warning = FALSE}
FinalDataTemp$qual7 <- relevel(FinalDataTemp$qual7, ref = "No qualifications")
FinalDataTemp$EthGrpGB <- relevel(FinalDataTemp$EthGrpGB, ref = 1)
lm3 <- lm(bmival ~ age + Sex + qual7 + NumCh118 +cigst2 +Saturatedfattyacidsg + FreeSugarsg + EthGrpGB + eqvinc, data = FinalDataTemp)
summary(lm3)
confint(lm3,  level = 0.95)
```

#### Interaction Effects 

```{r lm4, echo=TRUE, warning=FALSE}
lm4 <- lm(bmival ~ age + Sex + qual7 + NumCh118 +cigst2 +Saturatedfattyacidsg + FreeSugarsg + EthGrpGB + eqvinc + Saturatedfattyacidsg*EthGrpGB +FreeSugarsg *EthGrpGB + Saturatedfattyacidsg*eqvinc + FreeSugarsg * eqvinc + Saturatedfattyacidsg*EthGrpGB*eqvinc + FreeSugarsg*EthGrpGB*eqvinc, data = FinalDataTemp)
summary(lm4)
table(FinalDataTemp$EthGrpGB)
confint(lm4, level = 0.95)

# finding the mean income of each tertile 
data_mean <- aggregate(x = eqvinc ~ tertiles,              
                       data = FinalDataTemp,
                       FUN = mean)
```

#### Model diagnostics 

```{r diagnostics, echo=TRUE, warning=FALSE}
autoplot(lm4)
```

